On the presentation of orthogonal groups over polynomial rings

### 2019-07-24 dl-paper.tex
Reorganized introduction

\subsection{Relative Steinberg groups} \label{sec:quillen}
In this subsection we recall the definitions and basic facts pertaining to relative Steinberg groups as defined by J.-L.~Loday in~\cite{Lo78},
 where a theory of relative central extensions is developed and then applied to the usual (stable) Steinberg group $\St(R)$ from algebraic K-theory (see e.\,g.~\cite[Ch. III \S~5]{Kbook}).
The main goal of this subsection is to show that Loday's result can be applied to the unstable Steinberg groups, and that
 the resulting relative groups enjoy the same good properties as their stable counterparts.

%Some of the results of this subsection have been already mentioned in~\cite{S15} (cf. e.\,g. Corollaries 3--4).
%Nevertheless, we chose to reproduce them with greater amount of detail for the reason that~\cref{lem:prop41} depends on the unstable exact sequence of.

First of all, let us recall the definition of the category of (commutative) pairs $\catname{Pairs}$.
The objects of $\catname{Pairs}$ are pairs $(R, I)$, in which $R$ is a commutative ring and $I$ is an ideal of $R$.
By definition, a morphism $f \colon (R, I) \to (R', I')$ in the category of pairs corresponds to a ring map $f \colon R \to R'$ such that $f(I) \subseteq I'$.
Notice that the mapping $(R, I) \mapsto (R \to R/I)$ defines a functor from $\catname{Pairs}$ the morphism category $\catname{CRings}^\rightarrow$.

@@

Recall that {\it the double ring} $D_{R, I}$ of a pair $(R, I)$ is, by definition, the pullback ring $R \times_{R/I} R$.
In other words, it is the ring consisting of pairs of elements of $R$ congruent modulo $I$.
Denote by $p_0, p_1, \Delta$ the two obvious projections and the diagonal map $\xymatrix{D_{R,I} \ar@<1pt>[r] \ar@<-2.5pt>[r] & \ar@<-4pt>[l] R}$.

Let $S \colon \catname{CRings} \to \catname{Groups}$ be a functor.
Set $G_i = \Ker(S(p_i))$ and define {\it Loday's relativization} $S(R, I)$ as $ G_0 / [G_0, G_1]$. The map $S(p_1)$ induces a natural transformation $S(R, I) \to S(R)$. We denote this map by $\mu = \mu_{R,I}$ and its kernel by $C_S(R, I)$.
\begin{equation} \label{LodayRelativization}
 \xymatrix{ 1 \ar[r] & C_S(R, I) \ar[r] & S(R, I) \ar^{\mu}[r] & S(R) \ar[r] & S(R/I) \ar[r] & 1}
\end{equation}

@@

Our next goal is to obtain a more concrete description of the group $C(\Phi, R, I)$.
We need to recall some additional notation and terminology.

@@

Let $\nu \colon N \twoheadrightarrow Q$ be a surjective group-theoretic map.
A {\it relative central extension of $\nu$} is, by definition,
a crossed module $(M, N, \mu)$ such that the cokernel of $\mu$ is $\nu$.
The mentioned groups and morphisms fit into the following diagram.
\begin{equation} \label{RelativeCentralExtension}
 \xymatrix{ 1 \ar[r] & L \ar[r]^{\ker(\mu)} & M \ar^{\mu}[r] & N \ar^{\nu}[r] & Q \ar[r] & 1} \end{equation}

A morphism $(M, \mu) \to (M', \mu')$ of two relative central extensions of $\nu$ is, by definition, an $N$-group homomorphism $f\colon M \to M'$ such that $\mu' f = \mu$. 
It turns out that the set  $Ext(Q, N; L)$  of isomoprhism classes of relative central extensions of $\nu$ by an abelian group $L$ can be classified by means of a certain cohomological invariant called {\it characteristic class}. More precisely, \cite[Th{\'e}or{\`e}me~1]{Lo78} asserts that there is a well-defined bijection $\xi \colon Ext(Q, N; L) \to \HH^3(Q, N; L)$.

For the rest of this subsection $S \xrightarrow{\varphi} P \subseteq G$ is a triple of group-valued functors on the category of commutative rings satisfying the following assumptions:
\begin{enumerate} [label=(A\arabic*)]
 \item \label{req:left-exact} $G(D_{R, I}) \cong G(R) \times_{G(R/I)} G(R)$.
 \item \label{req:subfunc} For every $R$ the group $P(R)$ is a perfect normal subgroup of $G(R)$.
 \item \label{req:uce} For every $R$ the map $ \varphi_R \colon S(R) \to P(R)$ is a universal central extension.
 %This means that $\varphi_R$ is a certain surjective and that its kernel is contained in the center of $S(R)$.
 \item \label{req:coeq} For every pair $(R, I)$ the coequalizer of $S(p_0), S(p_1)$ is precisely $S(R) \to S(R/I)$.
\end{enumerate}
\begin{comment}
\begin{example}
 The obvious triple of functors satisfying the above assumptions is as follows.
 Take the infinite general linear group functor $\mathrm{GL}(-)$ as $G$, its infinite elementary subfunctor $E(-)$ as $P$
  and the infinite Steinberg group functor $\St(-)$ as $S$. %TODO: Why these 4 propeties hold? What is a relative central extension.
\end{example}
\end{comment}
\begin{lemma}
 Suppose  for every pair $(R, I)$ the map $\mu \colon S(R, I) \to S(R)$ is the universal relative central extension of $\nu \colon S(R) \to S(R/I)$, i.\,e. the initial object of the category of relative central extensions of $\nu$. 
 Moreover, $C_S(R, I)$ is naturally isomorphic to the relative homology group $\HH_3(S(R), S(R/I), \ZZ)$.
\end{lemma}
\begin{proof}
From~\ref{req:left-exact} and $\varphi(G_i) \subseteq \Ker(G(p_i))$ we conclude that $G_0 \cap G_1 \subseteq \Ker(\varphi \colon S(D_{R, I}) \to G(D_{R, I}))$ therefore $G_0 \cap G_1$ is a central subgroup of $S(D_{R, I})$. From~\ref{req:uce} and~\cite[Ch.~III, Theorem~5.4]{Kbook} we conclude that $\HH_1(S(R), \ZZ) = \HH_2(S(R), \ZZ) = 0$ for all rings $R$.

Using~\cite[Proposition~6]{Lo78} we obtain that the extension $S(R, I) \to S(R)$ is a universal relative central extension of the coequalizer $\mathrm{coeq}(d_0, d_1)$, which coincides with $S(R) \to S(R/I)$ by~\ref{req:coeq}.

Now set $C = \HH_3(S(R/I), S(R), \ZZ)$, $L = C_S(R, I)$.
Since 
Recall from the proof of \cite[Th{\'e}or{\`e}me~2]{Lo78} that there is a map $\alpha \colon C \to L.$
Under t

of~. To see that this isomorphism is natural in $(R, I)$ notice that the map $ $
\end{proof}

Thanks to the functoriality of Quillen's plus-construction (see e.\,g.~\cite[Proposition~5.2.4]{Ro95}) one can define a functor 
%TODO: can also refer to Weibel's IV.1.5 provided Top is replaced with hTop which apparently does not change anything
$X_{G, P}(R) = BG(R)^+_{P(R)} \colon \catname{CRings} \to \catname{Top}$ and then define for $i \geq 1$ the corresponding K-group functors $\K_{i}(G, P, R)$ as the homotopy groups $\pi_i(X_{G, P}(R))$.

%TODO: Give the definition of $D_{R, I}$ and of the category of Pairs
\begin{prop}\label{characterization}
For $(G, P, R)$ as above the following statements are true.
\begin{enumerate}
 \item There are natural isomorphisms
  \[ \K_2(G, P, R) \cong \Ker(S(R) \to G(R)), \ \K_3(G, P, R) \cong \HH_3(S(R), \ZZ). \]
 \item 
 \item There is a natural exact sequence:
 \[ \xymatrix{\K_3(G, P, R) \ar[r] & \K_3(G, P, R/I) \ar@{->>}[r] & C(R, I).} \]
 \end{enumerate}   
\end{prop}
\begin{proof}
 The first claim follows from~\ref{req:subfunc} and~\ref{req:uce} using the standard properties of the plus-construction, see \cite[\S~IV.1]{Kbook}   (cf. Exercises~1.8--1.9 ibid.). %provided one verifies that all the intermediate isomorphisms are natural in $R$.

 The third claim now follows from the starting portion of the homology long exact sequence of the map $S(R) \to S(R/I)$ and the first two assertions.
\end{proof}

\subsection{Relative Steinberg groups}
Let us briefly recall the definition and basic properties of relative Steinberg groups.
We refer the reader to~\cite[Section~3]{S15} for a more detailed exposition.
By definition, the relative Steinberg group functor $\St(\Phi, R, I)$ is obtained from the functor $\St(\Phi, R)$
 via the relativization procedure described in~\cref{sec:quillen}.

\subsection{Some elementary lemmas}
\begin{comment}
Hall-Witt identity
\[ [[ y^{-1}, x], z] ^ {y^{-1}}  [[ z^{-1}, y], x] ^ {z^{-1}}  [[ x^{-1}, z], y] ^ {x^{-1}} = 1 \]
\[ [[ y^{-1}, x], z]  \cdot [[ z^{-1}, y], x] ^ {z^{-1}y} \cdot  [[ x^{-1}, z], y] ^ {x^{-1}y} = 1. \]
\[ [[ z^{-1}, y^{-1}], x] ^ {z^{-1}y^{-1}} \cdot  [[ x^{-1}, z], y^{-1}] ^ {x^{-1}y^{-1}} = [z, [ y, x]] . \]
\[  . \]
\end{comment}
Throughtout this section $A$ denotes an arbitary ring, and $I, J$ two its ideals, and $\Phi$ a simply laced root system.

@@

\section{First injectivity theorem}
If we denote the kernel of the natural transformation $\mu$ from the definition of $\St(\Phi, R, I)$ by $C(\Phi, R, I)$ we obtain a natural exact sequence:
\begin{equation} \label{relSteinberg-def}
 \xymatrix{ 1 \ar[r] & C(\Phi, R, I) \ar[r]^{j} & \St(\Phi, R, I) \ar[r]^{\mu} & \St(\Phi, R) \ar[r]^{\pi} & \St(\Phi, R/I) \ar[r] & 1.}
\end{equation}
The group $C(\Phi, R, I)$ is trivial when $I$ is a splitting ideal of $R$, i.\,e. when the canonical map $R \to R/I$ splits, see \cite[Lemma~8]{S15}.
The following result plays a key role in the proof of~\cref{thm41}.
\begin{lemma} \label{lem:prop41}
In the notation of~\cref{lem:lemma32} assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical map $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$.
\end{lemma}
\begin{proof}
  Consider the following commutative diagram with rows obtained from~\cref{characterization}:
\begin{equation*}\xymatrix{
 KO_3(2\ell, B) \ar[r] \ar[d] & KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 KO_3(2\ell, R) \ar[r]        & KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
In view of~\cref{lem:Panin-stability} the central vertical map $f'$ is precisely the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
Invoking~\cref{bass-ft} (or~\cite[Corollary~5.3]{Ho05}) we obtain that \[\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k).\]
By \cite[Lemma~2.2]{FRS12} the group $\GW_2^{[3]}(k)$ is trivial, therefore $f'$ (and hence $f$) are surjective.
\end{proof}

>>>>

\subsection{Relative Steinberg groups and unstable K-groups} \label{sec:quillen}
In this subsection we recall the definitions and basic facts pertaining to relative Steinberg groups as defined by J.-L.~Loday in~\cite{Lo78}, where a theory of relative central extensions was developed and then applied to the stable Steinberg group $\St(R)$ from algebraic K-theory. The main goal of this subsection is to show that Loday's results can be applied to unstable Steinberg groups, and that the resulting relative groups enjoy many good properties of their stable counterparts.
Some of the results of this subsection have been briefly mentioned in~\cite{S15} (cf. e.\,g. Corollaries 3--4).

Recall that the category of (commutative) pairs $\catname{Pairs}$ is defined as follows.
Its objects are pairs $(R, I)$, in which $R$ is a commutative ring and $I$ is an ideal of $R$. A morphism of pairs $f \colon (R, I) \to (R', I')$ is, by definition, a ring map $f \colon R \to R'$ such that $f(I) \subseteq I'$. Notice that the mapping $(R, I) \mapsto (R \to R/I)$ defines a functor from $\catname{Pairs}$ the morphism category $\catname{CRings}^\rightarrow$.

@@

Recall that {\it the double ring} $D_{R, I}$ of a pair $(R, I)$ is, by definition, the pullback ring $R \times_{R/I} R$. In other words, it is the ring consisting of pairs of elements of $R$ congruent modulo $I$. Denote by $p_0, p_1, \Delta$ the two obvious projections and the diagonal map $\xymatrix{D_{R,I} \ar@<1pt>[r] \ar@<-2.5pt>[r] & \ar@<-4pt>[l] R}$. It is clear that $p_0 \Delta = p_1 \Delta = id_{R}$.

Let $S \colon \catname{CRings} \to \catname{Groups}$ be a functor. Set $G_i = \Ker(S(p_i))$ and define {\it Loday's relativization} $S(R, I)$ as $ G_0 / [G_0, G_1]$. The map $S(p_1)$ induces a natural transformation $S(R, I) \to S(R)$. We denote this map by $\mu = \mu_{R,I}$ and its kernel by $C_S(R, I)$: \begin{equation} \label{LodayRelativization} \xymatrix{ 1 \ar[r] & C_S(R, I) \ar[r] & S(R, I) \ar^{\mu}[r] & S(R) \ar[r] & S(R/I) \ar[r] & 1.} \end{equation}

@@

Our next goal is to obtain a homological interpretation of the group $C(\Phi, R, I)$.
For doing this we need to recall some additional notation and terminology.

@@

Let $\nu \colon N \twoheadrightarrow Q$ be a surjective group-theoretic map.
A {\it relative central extension of $\nu$} is, by definition,
a crossed module $(M, N, \mu)$ such that the cokernel of $\mu$ is $\nu$:
\begin{equation} \label{RelativeCentralExtension}
 \xymatrix{ 1 \ar[r] & L \ar[r] & M \ar^{\mu}[r] & N \ar^{\nu}[r] & Q \ar[r] & 1} \end{equation}

A morphism $(M, \mu) \to (M', \mu')$ of two relative central extensions of $\nu$ is, by definition, an $N$-group homomorphism $f\colon M \to M'$ such that $\mu' f = \mu$. 
A relative central extension is said to be {\it universal} if it is an initial object of the category of relative central extensions of $\nu$.

It turns out that the set  $Ext(Q, N; L)$  of isomorphism classes of relative central extensions of $\nu$ by an abelian group $L$ can be classified by means of a certain cohomological invariant called {\it characteristic class}. More precisely, \cite[Th{\'e}or{\`e}me~1]{Lo78} asserts that there is a well-defined bijection $\xi \colon Ext(Q, N; L) \to \HH^3(Q, N; L)$.

For the rest of this subsection $S \xrightarrow{\pi} P \subseteq G$ is a triple of group-valued functors on the category of commutative rings satisfying the following assumptions:
\begin{enumerate} [label=(A\arabic*)]
 \item \label{req:left-exact} $G(D_{R, I}) \cong G(R) \times_{G(R/I)} G(R)$.
 \item \label{req:subfunc} $P(R)$ is a perfect normal subgroup of $G(R)$.
 \item \label{req:coeq} For every pair $(R, I)$ the coequalizer of $S(p_0), S(p_1)$ is precisely $S(R) \to S(R/I)$.
 \item \label{req:uce} The map $ \pi_R \colon S(R) \to P(R)$ is a universal central extension for all $R$. In particular, $\HH_1(S(R), \ZZ) = \HH_2(S(R), \ZZ) = 0$.
\end{enumerate}

\begin{lemma}\label{lem:relativeH3}
 For every pair $(R, I)$ the map $\mu \colon S(R, I) \to S(R)$ is a universal relative central extension of $\nu \colon S(R) \to S(R/I)$. The group $C_S(R, I)$ is naturally isomorphic to the relative homology group $\HH_3(S(R), S(R/I), \ZZ)$.
\end{lemma}
\begin{proof}
The action of $S(R)$ on $S(D_{R, I})$ given by ${}^g h = S(\Delta)(g) \cdot h \cdot S(\Delta)(g)^{-1}$ induces an action of $S(R)$ on $S(R, I)$. With respect to this action the map $\mu \colon S(R, I) \to S(R)$ from~\eqref{LodayRelativization} is an $S(R)$-map.
From~\ref{req:left-exact} and $\varphi(G_i) \subseteq \Ker(G(p_i))$ we obtain that 
$G_0 \cap G_1 \subseteq \Ker(\pi_{D_{R, I}})$ and therefore is a central subgroup of $S(D_{R,I})$ by~\ref{req:uce}. Thus, we have verified the assumptions of~\cite[Proposition~6]{Lo78} which asserts that the map $\mu$ is a universal relative central extension of the coequalizer $\nu = \mathrm{coeq}(d_0, d_1)$. Since $\nu$ coincides with $S(R) \to S(R/I)$ by~\ref{req:coeq}, we have completed the proof of the first assertion of the lemma.

Set $N = S(R)$, $Q = S(R/I)$, $C = \HH_3(Q, N; \ZZ)$. Recall from the proof of~\cite[Th{\'e}or{\`e}me~2]{Lo78} that to every relative central extension $(M, \mu)$ of $\nu$ with kernel $L$ one can associate a map of abelian groups $C \to L$. This map is obtained from the characteristic class $\xi(M, \mu)$ via the isomorphism $\HH^3(Q, N; L) \cong \mathrm{Hom}(C, L)$ of the universal coefficients theorem.

In the special case $M = S(R, I)$ this construction produces a map $C \to C_{S}(R, I)$ which is natural in $(R, I)$ by~\cite[Proposition~3]{Lo78}. This map is an isomorphism by~\cite[Th{\'e}or{\`e}me~2]{Lo78}. \end{proof}

For a triple $(S, P, G)$ we could define the unstable Quillen K-groups functors as $\K_{i}^{G, P}(R) = \pi_i(BG(R)^+_{P(R)})$. The functors with lower numbers admit the following description.
\begin{lemma}\label{lem:lowerKgroups} There are natural isomorphisms \begin{enumerate}
 \item $\K_1^{G,P}(R) \cong G(R) / P(R)$;
 \item $\K_2^{G,P}(R) \cong \Ker(S(R) \to G(R))$;
 \item $\K_3^{G,P}(R) \cong \HH_3(S(R), \ZZ).$ \end{enumerate} \end{lemma}
\begin{proof} The first claim is obvious, the second and the third claim follow from~\ref{req:subfunc} and~\ref{req:uce} using the standard properties of the plus-construction, see \cite[\S~IV.1]{Kbook}   (cf. Exercises~1.8--1.9 ibid.) \end{proof}

\subsection{Some elementary lemmas}
\begin{comment}
Hall-Witt identity
\[ [[ y^{-1}, x], z] ^ {y^{-1}}  [[ z^{-1}, y], x] ^ {z^{-1}}  [[ x^{-1}, z], y] ^ {x^{-1}} = 1 \]
\[ [[ y^{-1}, x], z]  \cdot [[ z^{-1}, y], x] ^ {z^{-1}y} \cdot  [[ x^{-1}, z], y] ^ {x^{-1}y} = 1. \]
\[ [[ z^{-1}, y^{-1}], x] ^ {z^{-1}y^{-1}} \cdot  [[ x^{-1}, z], y^{-1}] ^ {x^{-1}y^{-1}} = [z, [ y, x]] . \]
\[  . \]
\end{comment}
Throughtout this section $\Phi$ is a simply laced root system, $A$ denotes a commutative ring, and $I, J$ are two arbitrary ideals of $A$.

@@

\section{First injectivity theorem}

If we denote the kernel of the natural transformation $\mu$ from the definition of $\St(\Phi, R, I)$ by $C(\Phi, R, I)$ we obtain a natural exact sequence:
\begin{equation} \label{relSteinberg-def}
 \xymatrix{ 1 \ar[r] & C(\Phi, R, I) \ar[r]^{j} & \St(\Phi, R, I) \ar[r]^{\mu} & \St(\Phi, R) \ar[r]^{\pi} & \St(\Phi, R/I) \ar[r] & 1.}
\end{equation}
The group $C(\Phi, R, I)$ is trivial when $I$ is a splitting ideal of $R$, i.\,e. when the canonical map $R \to R/I$ splits, see \cite[Lemma~8]{S15}.
The following result plays a key role in the proof of~\cref{thm41}.
\begin{lemma} \label{lem:prop41}
In the notation of~\cref{lem:lemma32} assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical map $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$. \end{lemma}
\begin{proof}

There is a natural exact sequence:
 \[ \xymatrix{\K_3(G, P, R) \ar[r] & \K_3(G, P, R/I) \ar@{->>}[r] & C_S(R, I).} \]
Writing the starting portion of the homology long exact sequence for the map $S(R) \to S(R/I)$ 
and using the isomorphisms of~\cref{lem:relativeH3} and~\cref{lem:lowerKgroups} we obtain a commutative diagram
\begin{equation*}\xymatrix{
 KO_3(2\ell, B) \ar[r] \ar[d] & KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 KO_3(2\ell, R) \ar[r]        & KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
In view of~\cref{lem:Panin-stability} the map $f'$ is precisely the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
Invoking~\cref{bass-ft} (or~\cite[Corollary~5.3]{Ho05}) we obtain that \[\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k).\]
By \cite[Lemma~2.2]{FRS12} the group $\GW_2^{[3]}(k)$ is trivial, therefore $f'$ (and hence $f$) are surjective.
\end{proof}
